---
title: "p8130_final_project"
author: "Brooklynn McNeil"
date: "2024-11-25"
output: html_document
---
```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
library(survival)
#library(survminer)

knitr::opts_chunk$set(
  comment = '', fig.width = 8, fig.height = 6, out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Exploratory Analysis and Visualization

The dataset we will be working with is from breast cancer patients. We are interested in predicting the risk of death based on many predictors from societal to genetic makeup.

Let's read in the data and take a look.

```{r}
breastcancer_df = read_csv("Project_2_data.csv") |>
  janitor::clean_names() |>
  mutate(sixth_stage = factor(x6th_stage)) |>
  select(-x6th_stage) |>
  mutate(
    across(c(race, marital_status, t_stage, n_stage, differentiate, a_stage, 
             estrogen_status, progesterone_status, status, sixth_stage), factor),
    status = case_when(
    status == "Alive" ~ 0,
    status == "Dead" ~ 1
  ))

head(breastcancer_df)
```

We have 16 variables relating to these patietns with breast cancer. The dataset includes `r breastcancer_df |> filter(status == 0) |>count(status) |> pull(n)` individuals that are alive and `r breastcancer_df |> filter(status == 1) |>count(status) |> pull(n)` that are dead. The average age of patients is `r breastcancer_df |> pull(age) |> mean() |> round()`. The races that are considered are `r unique(breastcancer_df$race)` and the marital statuses considered are `r unique(breastcancer_df$marital_status)`.

Let's look at our variables graphically.

```{r}
# age distribution
breastcancer_df |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Distribution of Age of Diagnosis")

# marital status distribution
breastcancer_df |>
  ggplot(aes(x = marital_status, fill = marital_status)) +
  geom_bar() +
  labs(title = "Distribution of Marital Status",
       x = "marital status")

# race disbtribution
breastcancer_df |>
  ggplot(aes(x = race, fill = race)) +
  geom_bar() +
  labs(title = "Distribution of Marital Status",
       x = "race")

T.stage = 
  breastcancer_df |>
  ggplot(aes(x = t_stage, fill = t_stage)) +
  geom_bar() +
  labs(title = "Distribution of T Stage",
       x = "t stage")
N.stage = 
  breastcancer_df |>
  ggplot(aes(x = n_stage, fill = n_stage)) +
  geom_bar() +
  labs(title = "Distribution of N Stage",
       x = "n stage") 

sixth.stage = 
  breastcancer_df |>
  ggplot(aes(x = sixth_stage, fill = sixth_stage)) +
  geom_bar() +
  labs(title = "Distribution of 6th Stage",
       x = "6th stage")

T.stage + N.stage / sixth.stage
```

Now let's look at the information pertaining to the tumors.

```{r}
# tumor size
breastcancer_df |>
  ggplot(aes(x = tumor_size)) +
  geom_histogram() +
  labs(title = "Distribution of Tumor Size",
       x = "tumor size (mm)")

# estrogen status
estrog = 
  breastcancer_df |>
  ggplot(aes(x = estrogen_status, fill = estrogen_status)) +
  geom_bar() +
  labs(title = "Estrogen Status of Tumor",
       x = "estrogen status")

# progesterone status
prog = 
  breastcancer_df |>
  ggplot(aes(x = progesterone_status, fill = progesterone_status)) +
  geom_bar() +
  labs(title = "Progesterone Status of Tumor",
       x = "progesterone status")

estrog + prog
```

## Statistical Analysis

Now let's start to look at the relationships between the variables and their affect on risk of death.

```{r}
# fit a cox regression model

breastcancer_df |>
  coxph(Surv(survival_months,status) ~ age + race + marital_status, data = _) |>
    broom::tidy()
```

