---
title: "p8130_final_project"
author: "Brooklynn McNeil"
date: "2024-11-25"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
library(survival)
library(survminer)
library(modelr)
library(car)

knitr::opts_chunk$set(
  comment = '', fig.width = 8, fig.height = 6, out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Exploratory Analysis and Visualization

The dataset we will be working with is from breast cancer patients. We are interested in predicting the risk of death based on many predictors from societal to genetic makeup.

Let's read in the data and take a look.

```{r}
breastcancer_df = read_csv("Project_2_data.csv") |>
  janitor::clean_names() |>
  mutate(sixth_stage = factor(x6th_stage)) |>
  select(-x6th_stage) |>
  mutate(
    status = case_when(
    status == "Alive" ~ 0,
    status == "Dead" ~ 1),
    across(c(race, marital_status, t_stage, n_stage, differentiate, a_stage, 
             estrogen_status, progesterone_status, sixth_stage), factor))

head(breastcancer_df)
```

We have 16 variables relating to these patietns with breast cancer. The dataset includes `r breastcancer_df |> filter(status == 0) |>count(status) |> pull(n)` individuals that are alive and `r breastcancer_df |> filter(status == 1) |>count(status) |> pull(n)` that are dead. The average age of patients is `r breastcancer_df |> pull(age) |> mean() |> round()`. The races that are considered are `r unique(breastcancer_df$race)` and the marital statuses considered are `r unique(breastcancer_df$marital_status)`.

Let's look at our variables graphically.

```{r}
# age distribution
breastcancer_df |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Distribution of Age of Diagnosis")

# marital status distribution
breastcancer_df |>
  ggplot(aes(x = marital_status, fill = marital_status)) +
  geom_bar() +
  labs(title = "Distribution of Marital Status",
       x = "marital status")

# race disbtribution
breastcancer_df |>
  ggplot(aes(x = race, fill = race)) +
  geom_bar() +
  labs(title = "Distribution of Marital Status",
       x = "race")

T.stage = 
  breastcancer_df |>
  ggplot(aes(x = t_stage, fill = t_stage)) +
  geom_bar() +
  labs(title = "Distribution of T Stage",
       x = "t stage")
N.stage = 
  breastcancer_df |>
  ggplot(aes(x = n_stage, fill = n_stage)) +
  geom_bar() +
  labs(title = "Distribution of N Stage",
       x = "n stage") 

sixth.stage = 
  breastcancer_df |>
  ggplot(aes(x = sixth_stage, fill = sixth_stage)) +
  geom_bar() +
  labs(title = "Distribution of 6th Stage",
       x = "6th stage")

T.stage + N.stage / sixth.stage
```

It looks like most of

Now let's look at the information pertaining to the tumors.

```{r}
# tumor size
breastcancer_df |>
  ggplot(aes(x = tumor_size)) +
  geom_histogram() +
  labs(title = "Distribution of Tumor Size",
       x = "tumor size (mm)")

# estrogen status
estrog = 
  breastcancer_df |>
  ggplot(aes(x = estrogen_status, fill = estrogen_status)) +
  geom_bar() +
  labs(title = "Estrogen Status of Tumor",
       x = "estrogen status")

# progesterone status
prog = 
  breastcancer_df |>
  ggplot(aes(x = progesterone_status, fill = progesterone_status)) +
  geom_bar() +
  labs(title = "Progesterone Status of Tumor",
       x = "progesterone status")

estrog + prog
```

Finally, let's look at the distribution of the outcome and address linearity.

```{r}
breastcancer_df |>
  ggplot(aes(x = age, y = survival_months, color = factor(status))) +
  geom_point(alpha = 0.7) 
```

## Statistical Analysis

Now let's start to look at the relationships between the variables and their affect on risk of death. Let's fit a cox regression model for age, race and T.stage as predictors. This test also assumes that the risk is constant over time, so we need to validate this assumption with proportional hazards.

```{r}
# fit a cox regression model

cox_fit = breastcancer_df |>
  coxph(Surv(survival_months,status) ~ age + race, data = _) 

cox_fit |>
  broom::tidy() |>
  knitr::kable()
```

Let's check for the assumptions of the cox model.

1.  Proportional hazards assumption

    1.  the effect of the covariates is constant over time

    2.  use `cox.zph()` to test for constant proportions

    3.  use `plot(cox.zph())` to plot, but Cox Regression does not depend on linearity.

2.  Linear relationship between covariates and log hazards.

    1.  asses with scatterplot

3.  Independence of survival times

    1.  If clustering exists, use a frailty model or robust standard errors to account for dependency.

4.  No omitted confounders

    1.  assumes we have included all relevant covariates in the model.

5.  No multicollinearity.

    1.  There should not be high colinearity between covariates.

    2.  use `vif()` variance inflation factor \>5 indicates multicolinearity.

```{r}
# check if the assumptions are valid
cox.zph(cox_fit)

# Extract the linear predictor (log-hazards)
log_hazards <- predict(cox_fit, type = "lp")

# Create a Q-Q plot for log-hazards
qqnorm(log_hazards, main = "Q-Q Plot of Log-Hazards vs Normal Distribution")
qqline(log_hazards, col = "red")

# check for mutlicolinearity
vif(cox_fit)
```

```{r}
# Kaplan-Meier survival curve by t_stage
km_fit <- survfit(Surv(survival_months, status) ~ t_stage, data = breastcancer_df)

# Plot the survival curves
ggsurvplot(km_fit, 
           data = breastcancer_df,
           xlab = "Months", 
           ylab = "Survival Probability",
           title = "Kaplan-Meier Survival Curve by T Stage",
           legend.title = "T Stage")
```

Ok, here's the plan

- map the survival model across all of the individual variables
- rule out the unimportant variables
- start adding in the variables to test if that improves the model (refer to diagnostics lecture)
- test for an increase in adjusted R squared for the best model


Now let's start with mapping across all variables!!

```{r}
models_list = list(
  age_mod = coxph(Surv(survival_months, status) ~ age, data = breastcancer_df),
  race_mod = coxph(Surv(survival_months, status) ~ race, data = breastcancer_df),
  marital_mod = coxph(Surv(survival_months, status) ~ marital_status, data = breastcancer_df),
  t_stage_mod = coxph(Surv(survival_months, status) ~ t_stage, data = breastcancer_df),
  n_stage_mod = coxph(Surv(survival_months, status) ~ n_stage, data = breastcancer_df),
  sixth_stage_mod = coxph(Surv(survival_months, status) ~ sixth_stage, data = breastcancer_df),
  dif_mod = coxph(Surv(survival_months, status) ~ differentiate, data = breastcancer_df),
  tumor_mod = coxph(Surv(survival_months, status) ~ tumor_size, data = breastcancer_df),
  region_examined_mod = coxph(Surv(survival_months, status) ~ regional_node_examined, data = breastcancer_df),
  region_pos_mod = coxph(Surv(survival_months, status) ~ reginol_node_positive, data = breastcancer_df),
  a_stage_mod = coxph(Surv(survival_months, status) ~ a_stage, data = breastcancer_df),
  estrogen_mod = coxph(Surv(survival_months, status) ~ estrogen_status, data = breastcancer_df),
  progesterone_mod = coxph(Surv(survival_months, status) ~ progesterone_status, data = breastcancer_df),
  grade_mod = coxph(Surv(survival_months, status) ~ grade, data = breastcancer_df)
)

# Tidy the models and combine into one data frame
single_results <- map_dfr(models_list, broom::tidy, .id = "model")

# View combined results
knitr::kable(single_results)
```


## Validation
```{r}
cv_df = crossv_mc(breastcancer_df,10) 

cv_df = cv_df |>
  mutate(
    train = map(train, \(i) as.tibble(i)),
    test = map(test, \(i) as.tibble(i))
  ) |>
  mutate(
    age_mod  = map(train, \(df) coxph(Surv(survival_months,status) ~ age, data = df)),
    race_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ race, data = df)),
    marital_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ marital_status, data = df)),
    t.stage_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ t_stage, data = df)),
    n.stage_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ n_stage, data = df)),
    sixth.stage_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ sixth_stage, data = df)),
    dif_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ differentiate, data = df)),
    tumor_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ tumor_size, data = df)),
    region_examined_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ regional_node_examined, data = df)),
    region_pos_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ reginol_node_positive, data = df)),
    a.stage_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ a_stage, data = df)),
    estrogen_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ estrogen_status, data = df)),
    progesterone_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ progesterone_status, data = df)),
    grade_mod = map(train, \(df) coxph(Surv(survival_months,status) ~ grade, data = df)))
```

